---
title: "Preprocess_Script_Batch1-4"
output: html_document
date: "2026-03-05"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# load singleCellTK

```{r}
library(singleCellTK)
```



# Read in SCE1

```{r}
base_dir <- "/restricted/projectnb/camplab/home/jiaqi/CTE_Batch1/outs/per_sample_outs"

# This is the *parent* directory where each sample lives
cellRangerDirs <- base_dir

sampleDirs <- list(c("K1010_DLF","K1107_DLF","K1535_DLF","K1571_DLF","K1308_DLF","K1322_DLF","K1461_DLF","K0551_DLF","K1244_DLF","K0933_DLF","K1396_DLF","K1217_DLF","K0982_DLF","K0509_DLF","K1475_DLF","K1058_DLF"))

sampleNames <-c("K1010_DLF","K1107_DLF","K1535_DLF","K1571_DLF","K1308_DLF","K1322_DLF","K1461_DLF","K0551_DLF","K1244_DLF","K0933_DLF","K1396_DLF","K1217_DLF","K0982_DLF","K0509_DLF","K1475_DLF","K1058_DLF")

sampleDirs

```


```{r}
# 2) import into a single SCE ----------------------------------------------

sce1 <- importCellRanger(
  cellRangerDirs = cellRangerDirs,
  sampleDirs      = sampleDirs,
  sampleNames     = sampleNames,
  cellRangerOuts = "count/sample_filtered_feature_bc_matrix/",
  class          = "Matrix"
)

```

```{r}
print(sce1)
```

```{r}
table(colData(sce1)$sample)
```

```{r}
# Get list of mitochondrial genes
sce1 <- importMitoGeneSet(sce1, reference = "human", id = "ensembl",
                         by = "rownames", collectionName = "mito")
```



# READ in SCE2

```{r}
base_dir <- "/restricted/projectnb/camplab/home/jiaqi/CTE_Batch2/outs/per_sample_outs"

# This is the *parent* directory where each sample lives
cellRangerDirs <- base_dir

sampleDirs <- list(c("K1308_EC","K0982_EC","K1475_EC","K1571_EC","K1322_EC", "K1058_EC", "K1535_EC", "K0551_EC","K1087_EC","K1010_EC","K1461_EC","K1217_EC","K1457_EC","K1396_EC","K0933_EC","K1244_EC"))

sampleNames <-c("K1308_EC","K0982_EC","K1475_EC","K1571_EC","K1322_EC", "K1058_EC", "K1535_EC", "K0551_EC","K1087_EC","K1010_EC","K1461_EC","K1217_EC","K1457_EC","K1396_EC","K0933_EC","K1244_EC")

sampleDirs

```


```{r}
# 2) import into a single SCE ----------------------------------------------

sce2 <- importCellRanger(
  cellRangerDirs = cellRangerDirs,
  sampleDirs      = sampleDirs,
  sampleNames     = sampleNames,
  cellRangerOuts = "count/sample_filtered_feature_bc_matrix/",
  class          = "Matrix"
)

```

```{r}
print(sce2)

```

```{r}
table(colData(sce2)$sample)
```

```{r}
# Get list of mitochondrial genes
sce2 <- importMitoGeneSet(sce2, reference = "human", id = "ensembl",
                         by = "rownames", collectionName = "mito")
```



# READ in SCE3: drop 1 low-count sample "K1396_CALC"

```{r}
base_dir <- "/restricted/projectnb/camplab/home/jiaqi/CTE_Batch3/outs/per_sample_outs"

# This is the *parent* directory where each sample lives
cellRangerDirs <- base_dir

sampleDirs <- list(c("K1475_CALC", "K1308_CALC", "K1010_CALC", "K1244_CALC", "K0982_CALC", "K0933_CALC", "K1058_CALC", "K1535_CALC", "K0551_CALC", "K1571_CALC", "K1461_CALC", "K1322_CALC", "K1217_CALC", "K0509_CALC", "K1107_CALC"))

sampleNames <-c("K1475_CALC", "K1308_CALC", "K1010_CALC", "K1244_CALC", "K0982_CALC", "K0933_CALC", "K1058_CALC", "K1535_CALC", "K0551_CALC", "K1571_CALC", "K1461_CALC", "K1322_CALC", "K1217_CALC", "K0509_CALC", "K1107_CALC")

sampleDirs

```

```{r}
# 2) import into a single SCE ----------------------------------------------

sce3 <- importCellRanger(
  cellRangerDirs = cellRangerDirs,
  sampleDirs      = sampleDirs,
  sampleNames     = sampleNames,
  cellRangerOuts = "count/sample_filtered_feature_bc_matrix/",
  class          = "Matrix"
)

```

```{r}
print(sce3)

```

```{r}
table(colData(sce3)$sample)
```

```{r}
# Get list of mitochondrial genes
sce3 <- importMitoGeneSet(sce3, reference = "human", id = "ensembl",
                         by = "rownames", collectionName = "mito")
```



# READ in SCE4: drop 1 low-count sample "K1331_DLF"

```{r}
base_dir <- "/restricted/projectnb/camplab/home/jiaqi/CTE_Batch4/outs/per_sample_outs"

# This is the *parent* directory where each sample lives
cellRangerDirs <- base_dir

sampleDirs <- list(c("K1218_DLF","K1650_DLF","K1121_DLF","SLI170_DLF","K1652_DLF","K0127_DLF","K0419_DLF","K0418_DLF","K1087_DLF","K1175_DLF","SLI111_DLF","K1573_DLF","K0965_DLF","K1565_DLF","K0483_DLF"))

sampleNames <-c("K1218_DLF","K1650_DLF","K1121_DLF","SLI170_DLF","K1652_DLF","K0127_DLF","K0419_DLF","K0418_DLF","K1087_DLF","K1175_DLF",
  "SLI111_DLF","K1573_DLF","K0965_DLF","K1565_DLF","K0483_DLF")

sampleDirs

```


```{r}
# 2) import into a single SCE ----------------------------------------------

sce4 <- importCellRanger(
  cellRangerDirs = cellRangerDirs,
  sampleDirs      = sampleDirs,
  sampleNames     = sampleNames,
  cellRangerOuts = "count/sample_filtered_feature_bc_matrix/",
  class          = "Matrix"
)

```


```{r}
print(sce4)

```

```{r}
table(colData(sce4)$sample)
```

```{r}
# Get list of mitochondrial genes
sce4 <- importMitoGeneSet(sce4, reference = "human", id = "ensembl",
                         by = "rownames", collectionName = "mito")
```



# Combine the 4 sces into 1 sce object

```{r}
# 1) put the objects in a list
sce_list <- list(b1 = sce1, b2 = sce2, b3 = sce3, b4 = sce4)

# 2) keep only the safe rowData columns
by_r <- c("feature_ID", "feature_name", "feature_type")

# add an explicit batch column
sce_list$b1$batch <- "batch1"
sce_list$b2$batch <- "batch2"
sce_list$b3$batch <- "batch3"
sce_list$b4$batch <- "batch4"

# 3) now combine 
sce_all <- combineSCE(
  sceList  = sce_list,
  by.r     = by_r,
  by.c     = NULL,        
  combined = TRUE
)

```


```{r}
table(sce_all$batch)
```


```{r}
table(sce_all$sample)
```


```{r}
print(sce_all)
```


# Run QC

```{r}
sce_all <- runCellQC(sce_all, sample = NULL,
                 algorithms = c("QCMetrics", "scDblFinder", "decontX"),
                 collectionName = "mito",
                 geneSetListLocation = "rownames")

```

# RUN QC UMAP

```{r}
sce_all <- getUMAP(inSCE = sce_all,  useReducedDim = NULL, useAssay = "counts", reducedDimName = "QC_UMAP", seed = 12345)

```


# Plot QC Result

```{r}

plotRunPerCellQCResults(sce_all)

```


# Plot doublets

```{r}

plotScDblFinderResults(sce_all, reducedDimName = "QC_UMAP")

```


#Plot contamination

```{r}

plotDecontXResults(sce_all, reducedDimName = "QC_UMAP")

```


#Generate QC result Report

```{r}
reportCellQC(sce_all)
```



#Filtering

```{r}
sce_all <- subsetSCECols(sce_all, colData = c("total > 600", "detected > 500",
                                      "mito_percent < 5"))

# See number of cells after filtering
ncol(sce_all)

```


```{r}
table(sce_all$batch)
```


```{r}
table(sce_all$sample)
```


```{r}
print(sce_all)
```


```{r}
rownames(sce_all) <- rowData(sce_all)$feature_name
sym <- rowData(sce_all)$feature_name
rownames(sce_all) <- make.unique(sym) 

```


# Normalization

```{r}
#normalize data, use decontx
sce_all <- runSeuratNormalizeData(inSCE = sce_all, useAssay = "decontXcounts", normAssayName = "seuratNormData")

#scale data
sce_all <- runSeuratScaleData(inSCE = sce_all, useAssay = "seuratNormData", scaledAssayName = "seuratScaledData")

```


# Variable Feature Selection

```{r}
#For each gene (row), it computes: its mean expression across cells; its variance, then “standardizes” that variance after modeling the mean–variance relationship. These statistics are stored in rowData(sce) (and/or internally for HVG).

sce_all <- runSeuratFindHVG(
  sce_all,
  useAssay = "counts",
  method = "vst"
)

print(getTopHVG(inSCE = sce_all, method = "vst", hvgNumber = 10))

#Plot mean vs standardized variance
#X-axis (Mean) – average expression of that gene across cells, on the transformed scale that the VST method uses (can be negative because of centering).
#Y-axis (Standardized Variance) – how much more variable the gene is than expected given its mean (after modeling the mean–variance trend).
#Red points = the genes selected as HVGs (e.g. top 2,000 by standardized variance

plotSeuratHVG(sce_all, labelPoints = 3)

```


# Dimensionality Reduction

```{r}

sce_all <- runSeuratPCA(inSCE = sce_all, useAssay = "seuratScaledData", reducedDimName = "seuratPCA", nPCs = 20)
#You see a large dense cluster on the right and an arched “tail” to the left. That suggests there are at least 2–3 major structures in the data:
#one big group (probably a dominant cell type), plus other cells whose expression gradually differs along that arch.
plotSeuratReduction(inSCE = sce_all, useReduction = "pca")

#If QC / normalization / HVG selection were broken, you’d often see:a big fuzzy ball with no structure, or PCs dominated by a few weird outlier cells.

```


```{r}

#Use PCs up to the elbow, so here something like 10 PCs is reasonable.
plotSeuratElbow(inSCE = sce_all)

```


# UMAP

```{r}
#UMAP builds a k-NN graph in that PC space and then lays the cells out in 2D.
sce_all <- runSeuratUMAP(inSCE = sce_all, useReduction = "pca", reducedDimName = "seuratUMAP", dims = 15)
plotSeuratReduction(sce_all, "umap")

```


# Clustering

```{r}
sce_all <- runSeuratFindClusters(inSCE = sce_all, useAssay = "seuratScaledData", useReduction = "pca", dims = 15, algorithm = "louvain")
plotSeuratReduction(sce_all, "pca", showLegend = TRUE)

```


# Save result into RDS
```{r}
saveRDS(sce_all, "/restricted/projectnb/camplab/home/jiaqi/sce_1234_QC_clustered.rds")
```

